---
title: "Project 1(Plastic)"
author: "Tyler Kinney, Greyson Ballesteros"
format: 
  html:
    code-tools: true
    code-fold: true
    embed-resources: true
    theme: journal
    toc: true
    toc-depth: 3
    toc-title: "Contents"
editor: source
execute:
  echo: true
  error: false
  warning: false
  message: false
---

```{r}
#| label: packages-setup
library(tidyverse)
library(glue)
library(here)
library(jsonlite)
library(purrr)
library(ggplot2)
library(ggridges)
library(gt)
library(scales)
library(ggtext)
```

## Research Question 1

RQ: Has the types of plastics that are being used changed significantly from 2019 to 2020?

### Data Context

```{r}
#| label: load_data
tuesdata <- tidytuesdayR::tt_load('2021-01-26')
plastics <- tuesdata$plastics
```

The data comes from Break Free from Plastic (BFFP) courtesy of Sarah Sauve. The observational unit is one cleaning project. The data contains variables relating to types of plastic, where the plastic comes from, and country. The columns in the data have been assigned certain classes like (double, character). They also cleaned the names of the columns using the janitor package. They also took two data sets from 2019 and 2020 and bound the rows together. The audit was completed in October of 2020, implying that the data was collected during that audit. The data was collected as part of a bigger project to understand the sources of plastic in different cities.

### Summary Tables

```{r}
#| label: Summary 1 (non-modified, iteration, custom function)
#| fig-alt: "Table displaying proportions of plastic type compared to the total in 2019 and 2020 and the change in those proportions between the years"
#| fig-cap: "Figure 1 shows the changes in proportions of each plastic type used from 2019 to 2020, with pp showing about a 14% increase and other showing about a 23% decrease."

plastic_types <- c("hdpe", "ldpe", "pet", "pp", "ps", "pvc", "o")

Prop_change <- function(type) {
  plastics |>
    filter(parent_company != "Grand Total") |>
    group_by(year) |>
    summarize(
      total_type = sum(across(type), na.rm = TRUE),
      total_all = sum(grand_total, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(prop = total_type / total_all) |>
    arrange(year) |>
    summarize(
      plastic_type = type,
      prop_2019 = prop[year == 2019],
      prop_2020 = prop[year == 2020],
      change = prop_2020 - prop_2019
    )
}

map_dfr(plastic_types, Prop_change) |>
  arrange(change) |>
  gt() |>
  tab_header(title = md("**Proportions of Each Plastic Type**"),
             subtitle = "and Changes in Those proportions (2020-2019)") |>
  fmt_percent(columns = c(`change`, prop_2019, prop_2020),
              decimals = 2) |>
  cols_label(plastic_type = "Type",
             `prop_2019` = "2019",
             `prop_2020` = "2020",
             change = "Change") |>
  tab_options(heading.background.color = "#EB6864",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
            color = "#333333",
            weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  data_color(columns = change,
             colors = col_numeric(
             palette = c("lightgreen", "white", "#D95C5C"),
             domain = NULL))
  
```

```{r}
#| label: Summary 2 (non-modified)
#| fig-alt: "Table displaying year over year increase in plastic type usage in 2019 and 2020."
#| fig-cap: "Figure 2 shows how plastic types increased from 2019 to 2020, with pvc showing the most increase with about 666% and other showing the most decrease with about 42%."

plastics |>
  filter(parent_company != "Grand Total") |>
  pivot_longer(cols = c(hdpe, ldpe, o, pet, pp, ps, pvc, grand_total),
               names_to = "plastic_type",
               values_to = "amount") |>
  group_by(plastic_type, year) |>
  summarize(total = sum(amount, na.rm = TRUE), .groups = "drop") |>
  arrange(plastic_type, year) |>
  group_by(plastic_type) |>
  mutate(`2019_2020_YoY` = (total - lag(total)) / lag(total)) |>
  filter(!is.na(`2019_2020_YoY`)) |>
  select(-total, -year) |>
  arrange(`2019_2020_YoY`) |>
  ungroup() |>
  gt() |>
  tab_header(title = md("**Year Over Year Increase in Types of Plastic Use**"),
             subtitle = "Calculated from 2020") |>
  fmt_percent(columns = `2019_2020_YoY`,
              decimals = 2) |>
  cols_label(plastic_type = "Type",
             `2019_2020_YoY` = "YoY Increase") |>
  tab_options(heading.background.color = "#EB6864",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
            color = "#333333",
            weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  data_color(columns = `2019_2020_YoY`,
             colors = col_numeric(
             palette = c("lightgreen", "white", "#D95C5C"),
             domain = NULL))

```

### Data Visualizations

```{r}
#| label: Visual 1
#| fig-alt: "Segmented bargraph displaying the distribution of plastic types in 2019 and 2020"
#| fig-cap: "Figure 3 shows the proportion of plastic types used per year from 2019 to 2020, with the graph showing that pp increased the most in usage and other decreased the most in usage."

journal_colors <- c("#FFFFFF", "#F8D6D6", "#F1ADAD", "#EA8585", "#E35C5C", "#DE4C4C", "#D95C5C")

plastics |>
  pivot_longer(cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
               names_to = "plastic_type",
               values_to = "plastic_amount") |>
  filter(parent_company != "Grand Total",
         !is.na(plastic_amount)) |>
  group_by(year, plastic_type) |>
  summarize(total = sum(plastic_amount, na.rm = TRUE),
            .groups = "drop") |>
  ggplot(aes(x = factor(year), y = total, fill = plastic_type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "Proportion of Plastic Types by Year",
       x = "Year",
       y = "Proportion",
       fill = "Plastic Type") +
  theme_light() +
  scale_fill_manual(values = journal_colors)
```

```{r}
#| label: Visual 2
#| fig-alt: "Slope graph showing changes in Proportions of each plastic type used from 2019 to 2020"
#| fig-cap: "Figure 4 shows the changes in proportion of plastic type usage per year from 2019 to 2020, with pp increasing the most and other decreasing the most."

plastics |>
  pivot_longer(cols = hdpe:pvc,
               names_to = "type",
               values_to = "count") |>
  filter(parent_company != "Grand Total",
         !is.na(count)) |>
  group_by(year, type) |>
  summarize(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  group_by(year) |>
  mutate(prop = total / sum(total)) |>
  ungroup() |>
  select(year, type, prop) |>
  pivot_wider(names_from = year,
              values_from = prop,
              names_prefix = "year_") |>
  mutate(change = year_2020 - year_2019) |>
  ggplot() +
  geom_segment(aes(x = year_2019, xend = year_2020,
               y = type, yend = type),
               color = "#333333",
               linewidth = 1) +
  geom_point(aes(x = year_2019, y = type), color = "grey", size = 3) +
  geom_point(aes(x = year_2020, y = type), color = "#EB6864", size = 3) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(title = "Change in Plastic Type Composition",
       subtitle = "(<span style='color:grey;'>2019</span> – <span style='color:#E45756;'>2020</span>)",
       x = "Proportion of total plastic",
       y = "Plastic type") +
  theme_light() +
  theme(plot.subtitle = element_markdown())
```

## Research Question 2

RQ: How does the distribution of plastic waste change with the different income levels of countries?

### Data Context

```{r}
# reading in csv, found here https://github.com/lukes/ISO-3166-Countries-with-Regional-Codes/blob/master/all/all.csv
GDP_2019_2020 <- read.csv(here("GDP_data.csv"), skip = 4) |>
  select(Country.Name, X2019, X2020) |>
  rename(`2019` = X2019, `2020` = X2020) |>
  pivot_longer(
    cols = c(`2019`, `2020`),
    names_to = "year",
    values_to = "gdp") |>
  mutate(year = as.numeric(year)) |>
  rename(country = Country.Name) |>
  mutate(
    country = recode(
      country,
      "Cote d'Ivoire" = "Cote D_ivoire",
      "Ecuador" = "ECUADOR",
      "Hong Kong SAR, China" = "Hong Kong",
      "Korea, Rep." = "Korea",
      "Nigeria" = "NIGERIA",
      "Turkiye" = "Turkey",
      "United Kingdom" = "United Kingdom of Great Britain & Northern Ireland",
      "United States" = "United States of America",
      "Viet Nam" = "Vietnam"
    )
  )


plastic_summary <- plastics |>
  filter(country != "EMPTY") |>
  group_by(country, year) |>
  summarize(total_plastic = sum(grand_total, na.rm = TRUE), .groups = "drop")

# combine the data
combined_data <- plastic_summary |>
  inner_join(GDP_2019_2020, by = c("country", "year"))

# matching countries to csv in order to match codes
plastics <- plastics |>
  filter(country != "EMPTY") |>
  mutate(country = recode(
    country,
    "Taiwan_ Republic of China (ROC)" = "Taiwan, Province of China",
    "Cote D_ivoire" = "Côte d'Ivoire",
    "ECUADOR" = "Ecuador",
    "NIGERIA" = "Nigeria",
    "Netherlands" = "Netherlands, Kingdom of the",
    "Tanzania" = "Tanzania, United Republic of",
    "Turkey" = "Türkiye",
    "United Kingdom" = "United Kingdom of Great Britain and Northern Ireland",
    "Vietnam" = "Viet Nam",
    "Korea" = "Korea, Republic of",
    "United Kingdom of Great Britain & Northern Ireland" = "United Kingdom of Great Britain and Northern Ireland"
  ))

Country_Codes <- read.csv(here('Country-Codes.csv'),
                          colClasses = c("character", "character")) |>
  rename(country = name) |>
  select(country, country.code)

# join

plastics_w_code <- plastics |>
  left_join(Country_Codes, by = "country")

# take unique country codes

codes <- plastics_w_code |>
  distinct(country.code) |>
  pull(country.code)

# api call function

collect_api_data <- function(code) {
  base_url <- 'https://restcountries.com/v3.1/alpha/'
  
  ctry_url <- glue('{base_url}','{code}')
  
  ctry_data <- fromJSON(ctry_url)
  
  if(length(ctry_data) == 0){
    return(NA)
  }
  
  else{
    ctry_data |>
      pull(languages)
  }
} 

# iterate through unique codes

Languages <- map(codes, collect_api_data)

# given a list of data frames formatted weirdly, so pivoting

Language_pivot <- function(list_df){
  list_df |>
    pivot_longer(cols = everything(),
                 names_to = 'Code',
                 values_to = 'Language') |>
    select(Language)
}

# iterate through the list of data frames

Languages <- map(Languages, Language_pivot)

# combine the languages into one value

Combine_lang <- function(list_df){
  list_df |>
    summarize(all_languages = paste(Language, collapse = ', '))
}

# iterate through list of data frames

Languages <- map(Languages, Combine_lang)

# pull languages

collect_lang <- function(list_df){
  list_df |>
    pull(all_languages)
}

# create languages data frame with codes and languages

languages_df <- tibble(
  country.code = codes,
  all_languages = map(Languages, collect_lang))

# join all data (including GDP data from Check-in 2), and separating languages (this is our "meta" dataset)

Updated_Plastics <- plastics_w_code |>
  filter(parent_company != "Grand Total") |>
  inner_join(languages_df, by = 'country.code') |>
  inner_join(combined_data, by = c('country', 'year')) |>
  separate_wider_delim(cols = all_languages,
                       delim = ", ",
                       names = paste0("lang", 1:11),
                       too_few = "align_start") |> 
  mutate(wealth_group = factor(ntile(gdp, 5), labels = c("Low", "Lower-Middle", "Middle", "Upper-Middle", "High")),
         plastic_per_gdp = total_plastic / gdp)

```

In order to assemble Updated_Plastics, we used the following data:

The GDP data is the country official statistics from the National Statistical Organizations and/or Central Banks. The observational unit is a country. The data has been collected since 1960 all the way up to 2024. It was obtained via the National Accounts data files, the Organisation for Economic Co-operation and Development (OECD), staff estimates, and the World Bank (WB). The data was collected for many reasons, but mainly for understanding and managing a country's economy.

The API data was collected and published by the REST Countries API project from official sources such as the United Nations, World Bank, and national statistical agencies. The observational unit is a country. It doesn't represent a single collection date. The data was collected indirectly by various sources via national censuses, surveys, and administrative records. The data was collected to provide a general country-level information about demographics and economies.

The country code data was collected by Luke Duncalfe. The observational unit is one country. The last revision made on the UN Statistical data was on June 19, 2024, from a document last revised on June 6, 2024. It was merged from both Wikipedia and the UN M49 Standard country or area codes for statistical data use. We chose to collect this data in order to make the links for the API we were using.

### Summary Tables

```{r}
#| label: Summary 3 (modified, Custom Function, Iteration)
#| fig-alt: "Table displaying the differences in proportions of plastic waste based on income class organized from 2019 to 2020."
#| fig-cap: "Figure 5 shows the proportion of plastic waste expulsion per country's income class from 2019 to 2020, with the lower-middle and middle classes using the most plastic waste and the low class using the least plastic waste."

years <- c(2019, 2020)

plastic_proportion_one_year <- function(year_input) {
  Updated_Plastics |>
    filter(year == year_input) |>
    group_by(wealth_group) |>
    summarise(
      total_plastic = sum(total_plastic, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(
      global_total = sum(total_plastic, na.rm = TRUE),
      proportion = total_plastic / global_total,
      year = year_input
    ) |>
    arrange(wealth_group)
}

table_data <- Updated_Plastics |>
  distinct(year) |>
  pull(year) |>
  map_dfr(plastic_proportion_one_year)

table_data |> 
  group_by(year) |>
  select(-global_total) |>
  gt() |>
  fmt_percent(columns = proportion) |>
  fmt_number(columns = total_plastic, decimals = 0) |>
  tab_header(title = md("**Proportion of Plastic Waste per Income**"), subtitle = "By Year") |>
  cols_label(wealth_group = "Wealth Group", total_plastic = "Total Plastic", proportion = "Proportion") |>
  tab_options(row_group.background.color = "#EB6964",
              heading.background.color = "#EB6964",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
            color = "#333333",
            weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  data_color(columns = proportion,
             colors = col_numeric(
             palette = c("#D95C5C", "white", "lightgreen"),
             domain = NULL))

```

```{r}
#| label: Statistical Model for Data in Summary 3/Figure 5
#| fig-alt: "Table displaying the differences in proportions of plastic waste based on income class organized from 2019 to 2020."
#| fig-cap: "The chi-squared model we created shows how there was significant evidence of differences present in proportions of plastic waste based on income class from 2019 to 2020."

plastics_country <- Updated_Plastics |>
  group_by(country, year, gdp) |>   
  summarize(
    hdpe = sum(hdpe, na.rm = TRUE), 
    ldpe = sum(ldpe, na.rm = TRUE),
    o = sum(o, na.rm = TRUE),
    pet = sum(pet, na.rm = TRUE),
    pp = sum(pp, na.rm = TRUE),
    ps = sum(ps, na.rm = TRUE),
    pvc = sum(pvc, na.rm = TRUE),
    .groups = "drop"               
  )

plastics_counts <- Updated_Plastics |>
  mutate(
    wealth_group = factor(
      ntile(gdp, 5),
      labels = c("Low", "Lower-Middle", "Middle", "Upper-Middle", "High")
    ) 
  ) |>
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "plastic_amount"
  ) |>
  group_by(wealth_group, plastic_type) |>
  summarize(
    total_plastic = sum(plastic_amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = plastic_type,
    values_from = total_plastic
  )

plastics_matrix <- as.matrix(plastics_counts[, -1])
rownames(plastics_matrix) <- plastics_counts$wealth_group

chisq <- chisq.test(plastics_matrix)

tibble(
  statistic = chisq$statistic,
  parameter = chisq$parameter,
  p.value = chisq$p.value,
  method = chisq$method) |>
  mutate(p.value = if(p.value < 0.001){
    "< 0.001"
  }
  else{
    p.value
  }) |>
  select(method, parameter, statistic, p.value) |>
  gt() |>
  fmt_number(
    columns = c(statistic, parameter),
    decimals = 2
  ) |>
  cols_label(
    statistic = "Chi-square statistic",
    parameter = "Degrees of freedom",
    p.value = "p-value",
    method = "Test"
  ) |>
  tab_header(
    title = md("**Chi-square Test of Plastic Waste Type by Income Group**"),
    subtitle = "Testing whether plastic waste type distribution differs across income levels"
  ) |>
  tab_options(heading.background.color = "#EB6864",
              column_labels.background.color = "lightgrey") |>
  tab_style(style = cell_borders(sides = "bottom",
            color = "#333333",
            weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle")))
  
```
A chi-square test of independence was conducted to examine whether the distribution of plastic waste types differs across income groups. The results were statistically significant (χ²(24) = 126,217.89, p-value < 0.001), indicating that the distribution of plastic waste types differs across income groups.

```{r}
#| label: Summary 4 (modified)
#| fig-alt: "Table displaying the differences in proportions of each plastic waste type based on income class from 2019 to 2020."
#| fig-cap: "Figure 6 shows the distribution of proportions of plastic waste types across different income levels from 2019 to 2020, with other being used the most by the middle class and pvc being used the least by the low class."

plastic_type_distribution <- Updated_Plastics |>
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "plastic_share"
  ) |>
  mutate(
    plastic_amount = total_plastic * plastic_share
  ) |>
  group_by(wealth_group, plastic_type) |>
  summarise(
    total = sum(plastic_amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(wealth_group) |>
  mutate(
    proportion = total / sum(total)
  ) |>
  ungroup() |>
  arrange(wealth_group, desc(proportion))

plastic_type_distribution |>
  gt(groupname_col = "wealth_group") |>
  fmt_percent(columns = proportion) |>
  fmt_number(columns = total, decimals = 0) |>
  cols_label(plastic_type = "Plastic Type", total = "Total Plastic", proportion = "Proportion") |>
  tab_header(title = md("**Distribution of Plastic Waste Types by Income Level**"), subtitle = "Composition within each wealth group") |>
  tab_options(row_group.background.color = "#EB6964",
              heading.background.color = "#EB6964",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
            color = "#333333",
            weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  data_color(columns = proportion,
             colors = col_numeric(
             palette = c("#D95C5C", "white", "lightgreen"),
             domain = NULL))

```

### Data Visualizations

```{r}
#| label: Visual 3
#| fig-alt: "Ridgeline plot displaying the total plastic waste divided by wealth groups in different countries from 2019 to 2020"
#| fig-cap: "Figure 7 shows how the total plastic waste was divided by income class in different countries from 2019 to 2020, with the middle class making the most and the low and high classes making the least."

journal_colors <- c("#FFFFFF", "#F8D6D6", "#F1ADAD", "#EA8585", "#E35C5C", "#DE4C4C", "#D95C5C")

Updated_Plastics |>
  mutate(
    wealth_group = factor(
      ntile(gdp, 5),
      labels = c("Low", "Lower-Middle", "Middle", "Upper-Middle", "High")
    )
  ) |>
  ggplot(aes(x = total_plastic, y = wealth_group, fill = wealth_group)) +
  geom_density_ridges(alpha = 0.7) +
  scale_fill_manual(values = journal_colors) +
  labs(
    title = "Distribution of Total Plastic Waste by Wealth Group",
    x = "Total Plastic Waste",
    y = "Wealth Group",
    fill = "Wealth Group"
  ) +
  theme_minimal()

```

```{r}
#| label: Visual 4
#| fig-alt: "Segmented bar graph displaying the proportion of average plastic types divided by wealth groups in different countries from 2019 to 2020"
#| fig-cap: "Figure 8 shows the proportion of average plastic types used by different income levels from 2019 to 2020. Lower-middle uses a lot of pet, middle uses a lot of pp, and upper-middle uses a lot of other plastics."

plastics_country <- Updated_Plastics |>
  group_by(country, year, gdp) |>   
  summarize(
    hdpe = sum(hdpe, na.rm = TRUE), 
    ldpe = sum(ldpe, na.rm = TRUE),
    o = sum(o, na.rm = TRUE),
    pet = sum(pet, na.rm = TRUE),
    pp = sum(pp, na.rm = TRUE),
    ps = sum(ps, na.rm = TRUE),
    pvc = sum(pvc, na.rm = TRUE),
    .groups = "drop"               
  ) |>
  mutate(
    wealth_group = factor(
      ntile(gdp, 5),
      labels = c("Low", "Lower-Middle", "Middle", "Upper-Middle", "High")
    )
  )

journal_colors <- c("#FFFFFF", "#F8D6D6", "#F1ADAD", "#EA8585", "#E35C5C", "#DE4C4C", "#D95C5C")

plastics_country |>
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",      
    values_to = "plastic_amount"    
  ) |>
  group_by(wealth_group, plastic_type) |>
  summarize(
    avg_plastic = mean(plastic_amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = wealth_group, y = avg_plastic, fill = plastic_type)) +
  geom_col(position = "fill") +   
  scale_fill_manual(values = journal_colors) +
  labs(
    title = "Proportion of Average Plastic Types by Wealth Group",
    x = "Wealth Group",
    y = "Proportion",
    fill = "Plastic Type"
  ) +
  theme_minimal()

```
