---
title: "Plastics During COVID: A Comparative Analysis of Plastic Waste from 2019-2020"
author: "Tyler Kinney, Greyson Ballesteros"
format: 
  html:
    code-tools: true
    code-fold: true
    embed-resources: true
    theme: journal
    toc: true
    toc-depth: 3
    toc-title: "Contents"
editor: source
execute:
  echo: true
  error: false
  warning: false
  message: false
---

```{r}
#| label: packages-setup
#| echo: false
library(tidyverse)
library(glue)
library(here)
library(jsonlite)
library(purrr)
library(ggplot2)
library(ggridges)
library(gt)
library(scales)
library(ggtext)
```

The introduction of novel coronavirus became a defining moment in global history when it made its first public appearance in Wuhan, China. The epidemic spread like wildfire, and has altered the course of humanity to this day. As the virus took its toll, humans began exhausting its resources like never before; one of these resources, of course, was plastics. As students that survived through the peak of COVID-19, we were interested in how the usage of plastics changed from 2019 to 2020, a span of years that has become one of the most relevant turning points in our lives.

## Plastic Usage Year to Year

Has the types of plastics that are being used changed significantly from 2019 to 2020?

### Data Context

```{r}
#| label: load_data
tuesdata <- tidytuesdayR::tt_load('2021-01-26')
plastics <- tuesdata$plastics
```

The data comes from Break Free from Plastic (BFFP) courtesy of Sarah Sauve. The observational unit is one cleaning project. The data contains variables relating to types of plastic, where the plastic comes from, and country. The columns in the data have been assigned certain classes like "double" or "character". The creators of the data set also cleaned the names of the columns using the janitor package. They also took two data sets from 2019 and 2020 and bound the rows together. The audit was completed in October of 2020, implying that the data was collected during that time frame The data was collected as part of a bigger project to understand the sources of plastic in different cities.

### Analysis

To start, we wanted to create a table that showed proportional differences in countries' usage of different plastic types from 2019 to 2020. The plastics we investigated were hdpe (high density polyethylene), ldp (low density polyethylene), pet (polyester plastic), pp (polypropylene), ps (polystyrene), pvc (polyvinyl chloride), and o (other plastic).

```{r}
#| label: Summary 1 (non-modified, iteration, custom function)
#| fig-alt: "Table displaying proportions of plastic type compared to the total in 2019 and 2020 and the change in those proportions between the years"

plastic_types <- c("hdpe", "ldpe", "pet", "pp", "ps", "pvc", "o")

Prop_change <- function(type) {
  plastics |>
    filter(parent_company != "Grand Total") |>
    group_by(year) |>
    summarize(
      total_type = sum(across(type), na.rm = TRUE),
      total_all = sum(grand_total, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(prop = total_type / total_all) |>
    arrange(year) |>
    summarize(
      plastic_type = type,
      prop_2019 = prop[year == 2019],
      prop_2020 = prop[year == 2020],
      change = prop_2020 - prop_2019
    )
}

map_dfr(plastic_types, Prop_change) |>
  arrange(change) |>
  gt() |>
  tab_header(title = md("**Proportions of Each Plastic Type**"),
             subtitle = "Changes in Proportions (2019-2020)") |>
  fmt_percent(columns = c(`change`, prop_2019, prop_2020),
              decimals = 2) |>
  cols_label(plastic_type = "Type",
             `prop_2019` = "2019",
             `prop_2020` = "2020",
             change = "Change") |>
  tab_options(heading.background.color = "#EB6864",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_caption(caption = "Table 1") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
            color = "#333333",
            weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  data_color(columns = change,
             colors = col_numeric(
             palette = c("lightgreen", "white", "#D95C5C"),
             domain = NULL))
  
```

Table 1 shows the changes in proportions of each plastic type used from 2019 to 2020. We can see that pp shows about a 14% increase in usage and other plastics showing about a 22% decrease in usage between the two years.

```{r}
#| label: Visual 1
#| fig-alt: "Slope graph showing changes in Proportions of each plastic type used from 2019 to 2020"
#| fig-cap: "Figure 1 shows the changes in proportion of plastic type usage per year from 2019 to 2020, with pp increasing the most and other decreasing the most."

plastics |>
  pivot_longer(cols = hdpe:pvc,
               names_to = "type",
               values_to = "count") |>
  filter(parent_company != "Grand Total",
         !is.na(count)) |>
  group_by(year, type) |>
  summarize(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  group_by(year) |>
  mutate(prop = total / sum(total)) |>
  ungroup() |>
  select(year, type, prop) |>
  pivot_wider(names_from = year,
              values_from = prop,
              names_prefix = "year_") |>
  mutate(change = year_2020 - year_2019) |>
  ggplot() +
  geom_segment(aes(x = year_2019, xend = year_2020,
               y = type, yend = type),
               color = "#333333",
               linewidth = 1) +
  geom_point(aes(x = year_2019, y = type), color = "grey", size = 3) +
  geom_point(aes(x = year_2020, y = type), color = "#EB6864", size = 3) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(title = "Change in Plastic Type Composition",
       subtitle = "(<span style='color:grey;'>2019</span> – <span style='color:#E45756;'>2020</span>)",
       x = "Proportion of total plastic",
       y = "Plastic type") +
  theme_light() +
  theme(plot.subtitle = element_markdown())
```

Next, we wanted to investigate how countries' plastic type year over year usage changed from 2019 to 2020, and we analyzed the same types of plastics.

```{r}
#| label: Summary 2 (non-modified)
#| fig-alt: "Table displaying year over year increase in plastic type usage in 2019 and 2020."

plastics |>
  filter(parent_company != "Grand Total") |>
  pivot_longer(cols = c(hdpe, ldpe, o, pet, pp, ps, pvc, grand_total),
               names_to = "plastic_type",
               values_to = "amount") |>
  group_by(plastic_type, year) |>
  summarize(total = sum(amount, na.rm = TRUE), .groups = "drop") |>
  arrange(plastic_type, year) |>
  group_by(plastic_type) |>
  mutate(`2019_2020_YoY` = (total - lag(total)) / lag(total)) |>
  filter(!is.na(`2019_2020_YoY`)) |>
  select(-total, -year) |>
  arrange(`2019_2020_YoY`) |>
  ungroup() |>
  gt() |>
  tab_header(title = md("**Year Over Year Increase in Types of Plastic Use**"),
             subtitle = "Calculated from 2020") |>
  fmt_percent(columns = `2019_2020_YoY`,
              decimals = 2) |>
  cols_label(plastic_type = "Type",
             `2019_2020_YoY` = "YoY Increase") |>
  tab_options(heading.background.color = "#EB6864",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_caption(caption = "Table 2") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
            color = "#333333",
            weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  data_color(columns = `2019_2020_YoY`,
             colors = col_numeric(
             palette = c("lightgreen", "white", "#D95C5C"),
             domain = NULL))

```

Table 2 shows the differences in plastic type usage measured as a proportion from 2019 to 2020. We observe that pvc shows the most increase in usage with about 364% and other plastic shows the most decrease with about 49%.

```{r}
#| label: Visual 2
#| fig-alt: "Segmented bargraph displaying the distribution of plastic types in 2019 and 2020"
#| fig-cap: "Figure 2 shows the proportion of plastic types used per year from 2019 to 2020, with the graph showing that pp increased the most in usage and other decreased the most in usage."

journal_colors <- c("#FFFFFF", "#F8D6D6", "#F1ADAD", "#EA8585", "#E35C5C", "#DE4C4C", "#D95C5C")

plastics |>
  pivot_longer(cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
               names_to = "plastic_type",
               values_to = "plastic_amount") |>
  filter(parent_company != "Grand Total",
         !is.na(plastic_amount)) |>
  group_by(year, plastic_type) |>
  summarize(total = sum(plastic_amount, na.rm = TRUE),
            .groups = "drop") |>
  ggplot(aes(x = factor(year), y = total, fill = plastic_type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "Proportion of Plastic Types by Year",
       x = "Year",
       y = "Proportion",
       fill = "Plastic Type") +
  theme_light() +
  scale_fill_manual(values = journal_colors)

#### NEED TO CHANGE
```

## Plastic Usage by Income Level

How does the distribution of plastic waste change with the different income levels of countries?

### Data Context

```{r}
# reading in csv, found here https://github.com/lukes/ISO-3166-Countries-with-Regional-Codes/blob/master/all/all.csv
GDP_2019_2020 <- read.csv(here("GDP_data.csv"), skip = 4) |>
  select(Country.Name, X2019, X2020) |>
  rename(`2019` = X2019, `2020` = X2020) |>
  pivot_longer(
    cols = c(`2019`, `2020`),
    names_to = "year",
    values_to = "gdp") |>
  mutate(year = as.numeric(year)) |>
  rename(country = Country.Name) |>
  mutate(
    country = recode(
      country,
      "Cote d'Ivoire" = "Cote D_ivoire",
      "Ecuador" = "ECUADOR",
      "Hong Kong SAR, China" = "Hong Kong",
      "Korea, Rep." = "Korea",
      "Nigeria" = "NIGERIA",
      "Turkiye" = "Turkey",
      "United Kingdom" = "United Kingdom of Great Britain & Northern Ireland",
      "United States" = "United States of America",
      "Viet Nam" = "Vietnam"
    )
  )


plastic_summary <- plastics |>
  filter(country != "EMPTY") |>
  group_by(country, year) |>
  summarize(total_plastic = sum(grand_total, na.rm = TRUE), .groups = "drop")

# combine the data
combined_data <- plastic_summary |>
  inner_join(GDP_2019_2020, by = c("country", "year"))

# matching countries to csv in order to match codes
plastics <- plastics |>
  filter(country != "EMPTY") |>
  mutate(country = recode(
    country,
    "Taiwan_ Republic of China (ROC)" = "Taiwan, Province of China",
    "Cote D_ivoire" = "Côte d'Ivoire",
    "ECUADOR" = "Ecuador",
    "NIGERIA" = "Nigeria",
    "Netherlands" = "Netherlands, Kingdom of the",
    "Tanzania" = "Tanzania, United Republic of",
    "Turkey" = "Türkiye",
    "United Kingdom" = "United Kingdom of Great Britain and Northern Ireland",
    "Vietnam" = "Viet Nam",
    "Korea" = "Korea, Republic of",
    "United Kingdom of Great Britain & Northern Ireland" = "United Kingdom of Great Britain and Northern Ireland"
  ))

Country_Codes <- read.csv(here('Country-Codes.csv'),
                          colClasses = c("character", "character")) |>
  rename(country = name) |>
  select(country, country.code)

# join

plastics_w_code <- plastics |>
  left_join(Country_Codes, by = "country")

# take unique country codes

codes <- plastics_w_code |>
  distinct(country.code) |>
  pull(country.code)

# api call function

collect_api_data <- function(code) {
  base_url <- 'https://restcountries.com/v3.1/alpha/'
  
  ctry_url <- glue('{base_url}','{code}')
  
  ctry_data <- fromJSON(ctry_url)
  
  if(length(ctry_data) == 0){
    return(NA)
  }
  
  else{
    ctry_data |>
      pull(languages)
  }
} 

# iterate through unique codes

Languages <- map(codes, collect_api_data)

# given a list of data frames formatted weirdly, so pivoting

Language_pivot <- function(list_df){
  list_df |>
    pivot_longer(cols = everything(),
                 names_to = 'Code',
                 values_to = 'Language') |>
    select(Language)
}

# iterate through the list of data frames

Languages <- map(Languages, Language_pivot)

# combine the languages into one value

Combine_lang <- function(list_df){
  list_df |>
    summarize(all_languages = paste(Language, collapse = ', '))
}

# iterate through list of data frames

Languages <- map(Languages, Combine_lang)

# pull languages

collect_lang <- function(list_df){
  list_df |>
    pull(all_languages)
}

# create languages data frame with codes and languages

languages_df <- tibble(
  country.code = codes,
  all_languages = map(Languages, collect_lang))

# join all data (including GDP data from Check-in 2), and separating languages (this is our "meta" dataset)

Updated_Plastics <- plastics_w_code |>
  filter(parent_company != "Grand Total") |>
  inner_join(languages_df, by = 'country.code') |>
  inner_join(combined_data, by = c('country', 'year')) |>
  separate_wider_delim(cols = all_languages,
                       delim = ", ",
                       names = paste0("lang", 1:11),
                       too_few = "align_start") |> 
  mutate(wealth_group = factor(ntile(gdp, 5), labels = c("Low", "Lower-Middle", "Middle", "Upper-Middle", "High")),
         plastic_per_gdp = total_plastic / gdp)

### DELETE API RELATED DATA

```

In order to assemble our Updated_Plastics dataset, we used the following data:

The GDP data is the country official statistics from the National Statistical Organizations and/or Central Banks. The observational unit is a country. The data has been collected since 1960 all the way up to 2024. It was obtained via the National Accounts data files, the Organisation for Economic Co-operation and Development (OECD), staff estimates, and the World Bank (WB). The data was collected for many reasons, but mainly for understanding and managing a country's economy.

The country code data was collected by Luke Duncalfe. The observational unit is one country. The last revision made on the UN Statistical data was on June 19, 2024, from a document last revised on June 6, 2024. It was merged from both Wikipedia and the UN M49 Standard country or area codes for statistical data use. We chose to collect this data in order to make the links for the API we were using.

### Analysis

For this section, we wanted to dive deeper into the distribution of plastic type usage by country from 2019 to 2020, but broken down by the wealth group/income level of each country.

```{r}
#| label: Summary 3 (modified, Custom Function, Iteration)
#| fig-alt: "Table displaying the differences in proportions of plastic waste based on income class organized from 2019 to 2020."

years <- c(2019, 2020)

plastic_proportion_one_year <- function(year_input) {
  Updated_Plastics |>
    filter(year == year_input) |>
    group_by(wealth_group) |>
    summarise(
      total_plastic = sum(total_plastic, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(
      global_total = sum(total_plastic, na.rm = TRUE),
      proportion = total_plastic / global_total,
      year = year_input
    ) |>
    arrange(wealth_group)
}

table_data <- Updated_Plastics |>
  distinct(year) |>
  pull(year) |>
  map_dfr(plastic_proportion_one_year)

table_data |> 
  group_by(year) |>
  select(-global_total) |>
  gt() |>
  fmt_percent(columns = proportion) |>
  fmt_number(columns = total_plastic, decimals = 0) |>
  tab_header(title = md("**Proportion of Plastic Waste per Income**"), subtitle = "By Year") |>
  cols_label(wealth_group = "Wealth Group", total_plastic = "Total Plastic", proportion = "Proportion") |>
  tab_options(row_group.background.color = "#EB6964",
              heading.background.color = "#EB6964",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_caption(caption = "Table 3") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
            color = "#333333",
            weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  data_color(columns = proportion,
             colors = col_numeric(
             palette = c("#D95C5C", "white", "lightgreen"),
             domain = NULL))

```

Table 3 shows the proportion of plastic waste expulsion per country's income class from 2019 to 2020. We can see that according to the table, the lower-middle and middle classes using the most plastic waste, while the lower class using the least plastic waste.

```{r}
#| label: Statistical Model for Data in Summary 3/Figure 3
#| fig-alt: "Table displaying the differences in proportions of plastic waste based on income class organized from 2019 to 2020."

plastics_country <- Updated_Plastics |>
  group_by(country, year, gdp) |>   
  summarize(
    hdpe = sum(hdpe, na.rm = TRUE), 
    ldpe = sum(ldpe, na.rm = TRUE),
    o = sum(o, na.rm = TRUE),
    pet = sum(pet, na.rm = TRUE),
    pp = sum(pp, na.rm = TRUE),
    ps = sum(ps, na.rm = TRUE),
    pvc = sum(pvc, na.rm = TRUE),
    .groups = "drop"               
  )

plastics_counts <- Updated_Plastics |>
  mutate(
    wealth_group = factor(
      ntile(gdp, 5),
      labels = c("Low", "Lower-Middle", "Middle", "Upper-Middle", "High")
    ) 
  ) |>
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "plastic_amount"
  ) |>
  group_by(wealth_group, plastic_type) |>
  summarize(
    total_plastic = sum(plastic_amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = plastic_type,
    values_from = total_plastic
  )

plastics_matrix <- as.matrix(plastics_counts[, -1])
rownames(plastics_matrix) <- plastics_counts$wealth_group

chisq <- chisq.test(plastics_matrix)

tibble(
  statistic = chisq$statistic,
  parameter = chisq$parameter,
  p.value = chisq$p.value,
  method = chisq$method) |>
  mutate(p.value = if(p.value < 0.001){
    "< 0.001"
  }
  else{
    p.value
  }) |>
  select(method, parameter, statistic, p.value) |>
  gt() |>
  fmt_number(
    columns = c(statistic, parameter),
    decimals = 2
  ) |>
  cols_label(
    statistic = "Chi-square statistic",
    parameter = "Degrees of freedom",
    p.value = "p-value",
    method = "Test"
  ) |>
  tab_header(
    title = md("**Chi-square Test of Plastic Waste Type by Income Group**"),
    subtitle = "Testing whether plastic waste type distribution differs across income levels"
  ) |>
  tab_caption(caption = "Table 4") |>
  tab_options(heading.background.color = "#EB6864",
              column_labels.background.color = "lightgrey") |>
  tab_style(style = cell_borders(sides = "bottom",
            color = "#333333",
            weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle")))
  
```

Table 4 displays a chi-square test of independence that was conducted to examine whether the distribution of plastic waste types differs across income groups. The results were statistically significant (χ²(24) = 126,217.89, p-value \< 0.001), indicating that the distribution of plastic waste types differs across income groups.

```{r}
#| label: Visual 3
#| fig-alt: "Ridgeline plot displaying the total plastic waste divided by wealth groups in different countries from 2019 to 2020"
#| fig-cap: "Figure 3 shows how the total plastic waste was divided by income class in different countries from 2019 to 2020, with the middle class making the most and the low and high classes making the least."

journal_colors <- c("#FFFFFF", "#F8D6D6", "#F1ADAD", "#EA8585", "#E35C5C", "#DE4C4C", "#D95C5C")

Updated_Plastics |>
  mutate(
    wealth_group = factor(
      ntile(gdp, 5),
      labels = c("Low", "Lower-Middle", "Middle", "Upper-Middle", "High")
    )
  ) |>
  ggplot(aes(x = total_plastic, y = wealth_group, fill = wealth_group)) +
  geom_density_ridges(alpha = 0.7) +
  scale_fill_manual(values = journal_colors) +
  labs(
    title = "Distribution of Total Plastic Waste by Wealth Group",
    x = "Total Plastic Waste",
    y = "Wealth Group",
    fill = "Wealth Group"
  ) +
  theme_minimal()

```

We then wanted to explore how the distributions of plastic type usage changed between the two years, specifically how they fluctuated within the different wealth group/income levels.

```{r}
#| label: Summary 4 (modified)
#| fig-alt: "Table displaying the differences in proportions of each plastic waste type based on income class from 2019 to 2020."

plastic_type_distribution <- Updated_Plastics |>
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "plastic_share"
  ) |>
  mutate(
    plastic_amount = total_plastic * plastic_share
  ) |>
  group_by(wealth_group, plastic_type) |>
  summarise(
    total = sum(plastic_amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(wealth_group) |>
  mutate(
    proportion = total / sum(total)
  ) |>
  ungroup() |>
  arrange(wealth_group, desc(proportion))

plastic_type_distribution |>
  gt(groupname_col = "wealth_group") |>
  fmt_percent(columns = proportion) |>
  fmt_number(columns = total, decimals = 0) |>
  cols_label(plastic_type = "Plastic Type", total = "Total Plastic", proportion = "Proportion") |>
  tab_header(title = md("**Distribution of Plastic Waste Types by Income Level**"), subtitle = "Composition within each wealth group") |>
  tab_caption(caption = "Table 5") |>
  tab_options(row_group.background.color = "#EB6964",
              heading.background.color = "#EB6964",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
            color = "#333333",
            weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  data_color(columns = proportion,
             colors = col_numeric(
             palette = c("#D95C5C", "white", "lightgreen"),
             domain = NULL))

```

Table 5 shows the distribution of proportions of plastic waste types across different income levels from 2019 to 2020. Some observations we made were that other plastics were used the most by the middle class, and pvc was used the least by the lower class.

```{r}
#| label: Visual 4
#| fig-alt: "Segmented bar graph displaying the proportion of average plastic types divided by wealth groups in different countries from 2019 to 2020"
#| fig-cap: "Figure 4 shows the proportion of average plastic types used by different income levels from 2019 to 2020. Lower-middle uses a lot of pet, middle uses a lot of pp, and upper-middle uses a lot of other plastics."

plastics_country <- Updated_Plastics |>
  group_by(country, year, gdp) |>   
  summarize(
    hdpe = sum(hdpe, na.rm = TRUE), 
    ldpe = sum(ldpe, na.rm = TRUE),
    o = sum(o, na.rm = TRUE),
    pet = sum(pet, na.rm = TRUE),
    pp = sum(pp, na.rm = TRUE),
    ps = sum(ps, na.rm = TRUE),
    pvc = sum(pvc, na.rm = TRUE),
    .groups = "drop"               
  ) |>
  mutate(
    wealth_group = factor(
      ntile(gdp, 5),
      labels = c("Low", "Lower-Middle", "Middle", "Upper-Middle", "High")
    )
  )

journal_colors <- c("#FFFFFF", "#F8D6D6", "#F1ADAD", "#EA8585", "#E35C5C", "#DE4C4C", "#D95C5C")

plastics_country |>
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",      
    values_to = "plastic_amount"    
  ) |>
  group_by(wealth_group, plastic_type) |>
  summarize(
    avg_plastic = mean(plastic_amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = wealth_group, y = avg_plastic, fill = plastic_type)) +
  geom_col(position = "fill") +   
  scale_fill_manual(values = journal_colors) +
  labs(
    title = "Proportion of Average Plastic Types by Wealth Group",
    x = "Wealth Group",
    y = "Proportion",
    fill = "Plastic Type"
  ) +
  theme_minimal()

```

## Conclusion

Overall, our analysis shows that plastic usage patterns shifted noticeably between the years of the start of the COVID-19 pandemic. We found that polypropylene (pp) had the most proportional increase, whereas "other" plastics exhibited a substantial decline. Additionally, some plastic types such as pvc (polyvinyl chloride) experienced dramatic year-over-year increases, suggesting changes in consumption behaviors during the epidemic.

When examining plastic waste by country income level, we observed that middle and lower-middle income groups contributed the largest share of plastic waste, while low and high income groups contributed the least. The chi-square test confirmed that the distribution of plastic types differs significantly across income levels, indicating that economic factors play an important role in shaping plastic consumption patterns.

In conclusion, these findings suggest that from 2019 to 2020, global events like COVID-19 and underlying economic differences between countries influenced not only how much plastic was used, but also what types of plastics were the most prevalent.

